---
title: "An overview of AnthropMMD"
author: "Frédéric Santos, <frederic.santos[at]u-bordeaux.fr>"
output: 
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
    fig_width: 5
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{An overview of AnthropMMD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: biblio.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(digits = 3)
```

```{r include=FALSE}
library(AnthropMMD)
```

# Graphical user interface

`AnthropMMD` was primarily designed as an R-Shiny application [@Santos18], that can be launched with the following instruction:
```{r eval=FALSE, include=TRUE}
start_mmd()
```

Starting with the version 3.0.0, it can also be used from the command line, to make it convenient in a context of reproducible research.

# Using `AnthropMMD` from the command line

## Data formatting

The MMD is a dissimilarity measure among groups of individuals described by binary (presence-absence) traits [@Sjovold73; @Sjovold77; @Harris04]. The MMD formula only requires to know the sample sizes and relative trait frequencies within each group: providing the raw individual data is not mandatory. 

Usually, the user will have recorded the data in one of the the two following formats.

### Raw binary data
Most of the time, the data will be formatted as a classical dataframe with $n$ rows (one per individual) and $p+1$ columns (the first column must be a group indicator; the $p$ other columns correspond to the $p$ binary traits observed). Each trait has two possible values: $1$ if present, $0$ if absent (missing values are allowed). Rownames are optional but colnames (i.e., `header`s) are mandatory.

An artifical dataset, `toyMMD`, is available in the package to give an example of such a format:

```{r rows.print=8}
data(toyMMD)
head(toyMMD)
```

It is not necessary for the `Trait`s to be manually converted as factors prior to the analysis. `toyMMD` includes nine traits with many missing values, and the individuals belong to five groups:

```{r}
str(toyMMD)
```

If the data were recorded as a raw binary dataset, they **must** be converted into a table of relative frequencies prior to the MMD analysis. The R function `binary_to_table` with the argument `relative = TRUE` can perform this operation:

```{r}
tab <- binary_to_table(toyMMD, relative = TRUE)
tab
```

### Tables of sample sizes and absolute frequencies
Sometimes, in particular if the data were extracted from research articles rather than true exprimentation, the user will only know the sample sizes and absolute trait frequencies in each group. As previously stated, those data are perfectly sufficient to compute MMD values. Here is an example of such a table:
```{r}
data(absolute_freqs)
absolute_freqs
```

If there are $k$ groups in the data, the first $k$ rows (whose names must start with an `N_` prefix, followed by the group labels) indicates the sample sizes of each trait in each group, and the last $k$ rows indicates the *absolute* trait frequencies. Such a table can be directly imported by the user, but **must** be converted to *relative* trait frequencies prior to the analysis, for instance by using the function `table_relfreq`:
```{r}
tab <- table_relfreq(absolute_freqs)
tab
```

### Important remark
Due to rounding issues, it is clearly not advised for the user to submit directly a table of relative frequencies. If the absolute frequencies do not perfectly sum up to 1 for each trait, an error will be triggered.

A table of relative frequencies is the starting point of all subsequent analyses, but it should be *computed* from the data loaded in R; it should not be loaded itself. 

A quick summary:

* if you load a raw binary dataframe, convert it to a table of relative frequencies with the function `binary_to_table(relative = TRUE)` as shown above;
* if you load a table of absolute frequencies, convert it to a table of relative frequencies with the function `table_relfreq`.

## Trait selection
`AnthropMMD` proposes a built-in feature for trait selection, in order to discard the traits that could be observed on too few individuals, or that do not show enough variability among groups. The function `select_traits` allows such a selection according to several strategies [@Harris04; @Santos18].

For instance, with the following instruction, we can discard the traits that could be observed on at least 10 individuals *per group*, and that exhibit no significant variability in frequencies among groups according to Fisher's exact tests:
```{r}
tab_selected <- select_traits(tab, k = 10, strategy = "keepFisher")
tab_selected$filtered
```

`Trait1` (which could be observed on only 8 individuals in Group B), `Trait5` and `Trait8` (whose variation in frequencies was not significant among groups) were removed from the data.

## Compute the mean measure of divergence
Once the trait selection has been performed, the MMD can be computed with the `mmd` function. With the following instruction, the MMD is computed using Anscombe's angular transformation of trait frequencies:
```{r}
mmd.result <- mmd(tab_selected$filtered, angular = "Anscombe")
mmd.result
```

* The first component, `$MMDMatrix`, follows the presentation adopted in most research articles [@Donlon00]: the true MMD values are indicated above the diagonal, and their standard deviations are indicated below the diagonal.
* A MMD value can be considered as significant if it is greater than twice its standard deviation. Significant MMD values are indicated by a `*` in the component `$MMDSignif`.
* Finally, the component `$MMDSym` is a symmetrical matrix of MMD values, with a null diagonal. This matrix of dissimilarities can be used to perform a multidimensional scaling or a hierarchical clustering to visualize the distances among the groups.

## Graphical representations of MMD
### Multidimensional scaling
Although `mmd.result$MMDSym` can perfectly be passed to your favourite function to produce a MDS plot, `AnthropMMD` also proposes a built-in function for such a graphical representation: `plot_mmd`. 

The MDS can be computed using the classical `stats::cmdscale` function (and the produces a metric MDS), or several variants of MDS algorithms implemented in the R package `smacof`.

For instance, we plot here the MDS coordinates computed with one variant of SMACOF algorithms:
```{r}
par(cex = 0.8)
plot_mmd(data = mmd.result$MMDSym, method = "interval", 
         gof = TRUE, axes = TRUE, xlim = c(-1.2, 0.75))
```

The argument `gof = TRUE` displays goodness of fits statistics for the MDS conigurations directly on the plot.

### Hierarchical clustering
`AnthropMMD` does not propose a built-in function for hierarchical clustering, but such a plot can easily be obtained with the usual R functions.
```{r}
library(cluster)
par(cex = 0.8)
plot(agnes(mmd.result$MMDSym), which.plots = 2, main = "Dendrogram of MMD dissimilarities")
```

# References