---
title: "An overview of AnthropMMD"
author: "Frédéric Santos, <frederic.santos[at]u-bordeaux.fr>"
output: 
  html_document:
    number_sections: true
    theme: cerulean
vignette: >
  %\VignetteIndexEntry{intro_AnthropMMD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r include=FALSE}
library(AnthropMMD)
```

# Graphical user interface

`AnthropMMD` was primarily designed as an R-Shiny application, that can be launched with the following instruction:
```{r eval=FALSE, include=TRUE}
AnthropMMD::start_mmd()
```

Starting with the version 3.0.0, it can also be used from the command line, to make it useful in a context of reproducible research.

# Using `AnthropMMD` from the command line

## Data formatting

Two types of datasets can be used:

* **Raw (binary) data**: a classical dataframe with $n$ rows (one per individual) and $p+1$ columns (the first column is a group indicator; the $p$ other columns correspond to the $p$ binary traits observed). Each trait has two possible values: $1$ if present, $0$ if absent (missing values are allowed). An artifical dataset
* **Tables of frquencies**